---
title: "Caulerpa sampling in the Ria Formosa"
subtitle: "Sampling transects and points"
author: "Márcio Martins"
date: "26 Jan 2018"
output:
  github_document:
    dev: "png"
always_allow_html: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, message = FALSE, warning = FALSE, fig.align = "center")
library(knitr)
library(kableExtra)
library(tidyverse)
library(lubridate)
library(rgdal)
library(ggmap)
library(plotly)
library(zoo)

points_palette <- c("start" = "#4daf4a", "end" = "#984ea3", "sampling point" = "#e41a1c", "estimated point" = "#377eb8")
```

## Caulerpa prolifera sampling transects
To determine the spatial distribution of *Caulerpa prolifera*, divers swam transects across a canal in the Ria Formosa lagoon (South Portugal) and stopped at pre-determined intervals to collect samples.

- The track of the divers was recorded by GPS, which was on a buoy attached to divers
- GPS track is continuous between transects, including boat travel
- There are a total of 4 transects
- Sampling points were also determined by having a user look at footage of the dives, mark the times at which samples were taken and them crossing them with the GPS data. These points were then used to confirm if this automatic method is accurate

**GOALS:**  
1. To determine start and end point of each transect, isolate and extract them.  
2. Determine sampling points along the transect.  

### Isolating transects

Let's start by importing the tracks. File format is ".gpx". For convenience, I will convert this information into a tabular format. There are 4 also tracks on record. We are only interested in the last one.  
```{r track}
track <- readOGR("./GPS_tracks.gpx", layer = "track_points")

points <- tibble("trackID" = track@data$track_fid,   #Select variables we are interested in
                 "time"    = track@data$time,
                 "lat"     = track@coords[,2],
                 "long"    = track@coords[,1]) %>%
          filter(trackID == 4) %>%                   #Only keep track 4
          select(-trackID)                           #Remove trackID variable

#Time is imported as a factor, change to POSIXct date
points$time <- as.character(points$time) %>%
               ymd_hms()

head(points)
```

Now the information is in a tibble (a "modern take on data frames", in the words of the package creator). This format is very easy to manipulate, as well as plot using ggplot2.

Let's start by getting a glimpse of the full track:

```{r track_plot}
#Assign a number ID to the points, it will be used to identify points later
points$nr <- c(1:nrow(points))

map <- get_map(location = c(long = -7.819869, lat = 37.00763),
               color = "color",
               source = "google",
               maptype = "satellite",
               zoom = 14)
ggmap(map) +
  geom_path(data = points, aes(x = long, y = lat), size = 1) +
  lims(x = c(min(points$long), max(points$long)), y = c(37.00176, 37.01067)) +
  labs(x = "Longitude", y = "Latitude")
```

Using this information, we need to extract the starting and ending points of each transect. Visually, I can easily find all of the tracks. Since this is possible, I will create an interactive map to allow me to extract the ID number of these points.  

> Due to limitations in github .md files, this does not work on the github page.  

```{r interactive_track, include = FALSE}
#Let's select every other 10th point for plotting to ease the size of the map
points_simple <- points[seq(1,nrow(points),10),]

ggplotly(
  ggmap(map) +
  geom_point(data = points_simple, aes(x = long, y = lat, color = nr), size = 0.5) +
  scale_color_gradient2(low = "#a50026", mid = "#ffffbf", high = "#313695", midpoint = nrow(points)/2) +
  lims(x = c(min(points$long), max(points$long)), y = c(min(points$lat), max(points$lat))) +
  labs(x = "Longitude", y = "Latitude") +
  theme(legend.position = "none")
  )
```

Using this interactive map we can approximate the first and last points of each transect and obtain their ID:  
- Transect one:   Points 1 to 581  
- Transect two:   Points 2401 to 3811  
- Transect three: Points 5311 to 6601  
- Transect four: Points 8311 to 9571  
  
We can now extract the transects.  
```{r transects}
t1 <- points %>%
      filter(nr <= 581)
t1$transect <- 1

t2 <- points %>%
      filter(nr >= 2401 & nr <= 3811)
t2$transect <- 2

t3 <- points %>%
      filter(nr >= 5311 & nr <= 6601)
t3$transect <- 3

t4 <- points %>%
      filter(nr >= 8311 & nr <= 9571)
t4$transect <- 4

transects <- rbind(t1,t2,t3,t4) %>%
             mutate(transect = as.factor(transect))

ggmap(map) +
  geom_path(data = transects, aes(x = long, y = lat, color = transect, group = transect), size = 0.5) +
  scale_color_manual(values = c("#e41a1c", "#4daf4a", "#984ea3","#ff7f00")) +
  lims(x = c(min(points$long), max(points$long)), y = c(37.00176, 37.01067)) +
  labs(x = "Longitude", y = "Latitude") +
  theme(legend.position = "none")
```

The resulting transects seem accurate. However, results should be confirmed by checking starting and ending times for transects in the videos. To make this task easier, let's make a table with start/end times for the transects obtained here.

```{r time_table}
kable(transects %>%
      group_by(transect) %>%
      summarise(Start = min(time), End = max(time)),
  align = c("c","c","c"),
  format = "html") %>%
kable_styling(full_width = F)

```

### Sampling points
Now that we have the transects, let's see how accurate an automated methodology to determine sampling points would be. With one point per second, we have a pretty high resolution track for the transects. An interesting method might be to estimate diver's speeds along the transects to find stopping points.  

To do so, we will:  

- Calculate the total distance travelled from start of transect until current point  
- Estimate travel speed at each point  

The distance between any two points p1 and p2 is given by:
Distance <sub>p1 to p2</sub> = Square root ((X<sub>2</sub> - X<sub>1</sub>)<sup>2</sup> + (Y<sub>2</sub> - Y<sub>1</sub>)<sup>2</sup>)

The full distance travelled from the start of the transect to any given point is then simply the cumulative sum of distance. To this we will call Delta Distance.  

The diver speed at any given point can be calculated as the speed over the previous 10 seconds:  
Speed<sub>at time</sub> = (Delta Distance<sub>time - 10</sub> - Delta Distance<sub>time</sub>)/10  


Because we're only interested in how fast they are going relative to other points in the transect, we will leave distance units as degrees and will calculate speed as degrees per second.  

```{r speed}
transects <- transects %>% 
             group_by(transect) %>%
             mutate(distance = sqrt((lat - lag(lat))^2 + (long - lag(long))^2)) 

transects$distance[is.na(transects$distance)] <- 0

transects <- transects %>%
             group_by(transect) %>%
             mutate(distance = cumsum(distance - min(distance)))

transects <- transects %>%
             group_by(transect) %>%
             mutate(speed = (distance - lag(distance, n = 10, default = 0))/10)

ggplot(transects) +
  geom_line(aes(x = distance, y = speed)) +
  facet_wrap(~ transect, scales = "free") +
  labs(x = "Distance along the transect (degrees)", y = "Speed (degrees/second)") +
  theme_bw()
```

This looks interesting. This visualization shows too much variation. Smoothing (running average) the plots might make points where the divers stop easier. Another alternative is simply to plot the transects, with color based on speed. Because average speed was different amongst transects, let's increase contrast of the colors. To do so, I will standardize the speed within each individual transect (subtract mean and divide by sd). We lose information about speed differences amongst transects but keep information about the parts of a transect that were slow, compared to the rest of the same transect. 

```{r scaled_speed}
transects <- transects %>%
             group_by(transect) %>%
             mutate(scaled_speed = (speed - mean(speed))/sd(speed),
                    extremety = case_when(nr == min(nr) ~ "start",
                                          nr == max(nr) ~ "end",
                                          TRUE ~ "no"))

ggplot() +
  geom_path(data = transects,
            aes(x = long, y = lat, color = scaled_speed), 
            size = 2) +
  scale_color_gradient(high = "#ff0008", low = "#0061ff") +
  geom_point(data = transects %>% filter(extremety %in% c("start","end")),
             aes(x = long, y = lat, fill = extremety),
             size = 3,
             shape = 21) +
  scale_fill_manual(values = points_palette) + 
  facet_wrap(~ transect, scales = "free")+
  labs(x = "Longitude", y = "Latitude", color = "Scaled speed", fill = "") +
  theme_bw()
```

This is promising. Parts of slower movement are visible along the transect. Let's add even more contrast by dividing points based on quantiles. Sections with speed below the decided quantile will be colored blue. A bit of trial and error showed that 40th quantile is a good compromise for all transects.  

```{r speed_quantile}
transects <- transects %>%
             group_by(transect) %>%
             mutate(quant40_speed= ifelse(scaled_speed < quantile(scaled_speed, 0.40), "slow", "fast"))

ggplot() +
  geom_path(data = transects, aes(x = long, y = lat, color = quant40_speed, group = transect), size = 2) +
  scale_color_manual(values = c("#ff0008","#0061ff")) +
  facet_wrap(~ transect, scales = "free")+
  geom_point(data = transects %>% filter(extremety %in% c("start","end")),
             aes(x = long, y = lat, fill = extremety),
             size = 3,
             shape = 21) +
  scale_fill_manual(values = points_palette) + 
  labs(x = "Longitude", y = "Latitude", color = "Speed", fill = "") +
  theme_bw()
```

We're getting somewhere now.  

### Estimated vs real sampling points
Sampling points were manually determined using video footage of the dive to extract time at which sampling ocurred. The GPS data was then used to determine divers position at that specific time.  To see how accurate we were at extracting these sampling points, we can now plot automatic and manually determined points and see how close they are.


```{r points_plot}
#Function that identifies when the divers are stopped and gives those points an ID number
identify_points <- function(vector,interval){
  interval <- interval - 1
  point_id <- 1                       #initialize a number to identify sampling points
  ids <- vector(mode = "numeric")     #Vector to stopre sampling points id number
  for(i in 1:length(vector)){
    #If 10 points in a row are considered slow, return TRUE (divers are stoped)
    stopped <- all(c(vector[i:i+interval]) == "slow")
    
    ids <- ids %>%          #If diver is stoped, give point an ID, else NA
           c(ifelse(stopped,
           point_id,
           NA))
    
    point_id <- ifelse(stopped, point_id, point_id + 1) #If divers are moving, increase point ID counter. This does not create a continuous ID (there are gaps), but that is not necessary
  }
  return(ids)
}

transects_smooth <- transects %>%         #To plot smooth transects I will average every 10 points and plot a path
                    select(lat, long, nr, transect, extremety) %>%
                    group_by(transect) %>%
                    mutate(long = rollmean(long, 20, fill = NA, aling = "center"),
                           lat  = rollmean(lat, 20, fill = NA), aling = "center") %>%
                    filter(!is.na(lat)) %>%
                    .[seq(1, nrow(.), 20),]  %>% #Select every 10th point
                    ungroup()

sampling_points <- read_csv2("./sampling_points.csv") %>%  #Import the manually determined points
                   rename(transect = Transect, long = Long, lat = Lat) %>%
                   mutate(name = "sampling point")

estimated_points <- transects %>%
                    mutate(sample_points = identify_points(quant40_speed, 15)) %>%
                    filter(!is.na(sample_points)) %>%
                    group_by(transect, sample_points) %>%
                    summarise(lat = mean(lat), long = mean(long)) %>%
                    mutate(name = "estimated point") %>%
                    ungroup()

plot_points <- transects %>% ungroup %>%
               filter(extremety %in% c("start","end")) %>%
               select(transect, long, lat, name = extremety) 

plot_points <- plot_points %>%
               rbind(sampling_points %>% select(transect, long, lat, name))

plot_points <- plot_points %>%
               rbind(estimated_points %>% select(transect, long, lat, name))

ggplot() +
  geom_path(data = transects_smooth, aes(x = long, y = lat, group = transect), size = 1) +
  geom_point(data = plot_points, aes( x = long, y = lat, color = name, size = name)) +
  facet_wrap(~ transect, scales = "free") +
  scale_color_manual(values = points_palette) + 
  scale_size_manual(values = c("start" = 3, "end" = 3, "sampling point" = 2, "estimated point" = 1), guide = FALSE) +
  labs(x = "Longitude", y = "Latitude", color = "Speed", fill = "") +
  theme_bw()
```


This methodology seems to be useful to approximate sampling locations without manually going through video footage. One pattern is also obvious: it was much more accurate in the last 2 transects than the first ones.  There are also some points that were not detected at all. These issues might be explained by:  

- One of the divers was using a scuba scooter for the first time, and the first dive was done against the current. This made speed along the first transect fluctuate more than in other cases  
- In areas where there was no *Caulerpa prolifera*, the divers simply kept going, meaning there is no decrease in speed to detect  
- Part of transect one was not recorded in the GPS data (memory was full and it was overwritten)  

**Much of this can be fixed by setting some rules for field work:**  
- Change GPS to record points every 5 seconds, rather than 1 increases the time we can record 5-fold, while maintaining a reasonably high resolution (make sure to carry an extra GPS just in case)  
- Ensure divers ALWAYS stop a minimum of 30 seconds per sampling point, even if there is no algae to sample in the area  
- Record transect start and end time, so that transects can be extracted easily by filtering points between certain time periods

**The reliability of the processing can be improved by:**  
- Smoothing the data (a simple running average would likely be enough) to decrease noise in the data. This might not be as important if recording interval is changed from 1 second to 5, but it's worth looking into  
- As a check for accuracy of results, extract number of estimated sampling points and see if they match real number of samples (easy to do at this point)  

